---
title: "Course 5 - Week 2 Assignment"
author: "Noel Figuera"
date: "May 15, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

* Code for reading in the dataset and/or processing the data
* Histogram of the total number of steps taken each day
* Mean and median number of steps taken each day
* Time series plot of the average number of steps taken
* The 5-minute interval that, on average, contains the maximum number of steps
* Code to describe and show a strategy for imputing missing data
* Histogram of the total number of steps taken each day after missing values are imputed
* Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
* All of the R code needed to reproduce the results (numbers, plots, etc.) in the report


```{r }

library(ggplot2)
library(dplyr)

path <- 'activity.csv'
df <- read.csv(path,stringsAsFactors = FALSE)

# load the dataset

df$date <- as.Date(df$date,format="%Y-%m-%d")

# create a factor with two levels, weekday or weekend
df$weekday <- sapply(df$date, function(x) {
      if (weekdays(x) == "Sunday" || weekdays(x)=="Saturday") {
            return ('WEEKEND')
      } else {
            return ('WEEKDAY')
      }})

df$weekday <- as.factor(df$weekday)

# first analysis include all nas
df_na <- df[!is.na(df$steps),]

# total steps taken by day
s <- df_na %>% group_by(date) %>% summarise(records=n(), total_steps=sum(steps))

# create an histogram
ggplot(s, aes(total_steps)) + geom_histogram(bins=50)

# calculate mean and median
summary(s$total_steps)

# calculate the average per time interval
t <- df_na %>% group_by(interval) %>% summarise(records=n(), average_steps=mean(steps), total_steps=sum(steps))

# plot the lines using ggplot2
ggplot(t, aes(x=interval, y=average_steps)) + geom_line()

# which on average contains the max?
head(t[order(-t$average_steps),])
#835

# analyse number of complete cases vs with nas
x <- c(length(df[!is.na(df$steps),]$steps), length(df[is.na(df$steps),]$steps))
names(x) <- c("Complete Cases", "With NAs")
barplot(x)

# fill the na's with values using a simple strategy, substitute using the average of the 
# interval average
df_fill <- df
df_fill$steps[is.na(df$steps)] <- t$average_steps[match(df$interval,t$interval)][which(is.na(df$steps))]

# validate the results
x <- c(length(df_fill[!is.na(df_fill$steps),]$steps), length(df_fill[is.na(df_fill$steps),]$steps))
names(x) <- c("Complete Cases", "With NAs")
barplot(x)

##################################################################
#  We repeat the whole analysis for the dataset without NAs


# total steps taken by day
s2 <- df_fill %>% group_by(date) %>% summarise(records=n(), total_steps=sum(steps))

# create an histogram
ggplot(s2, aes(total_steps)) + geom_histogram(bins=50)

# calculate mean and median
summary(s2$total_steps)

# calculate the average per time interval
t2 <- df_fill %>% group_by(interval) %>% summarise(records=n(), average_steps=mean(steps), total_steps=sum(steps))

# plot the lines using ggplot2
ggplot(t2, aes(x=interval, y=average_steps)) + geom_line()

# which on average contains the max?
head(t2[order(-t2$average_steps),])

##################################################################
# Identify patterns on weekdays vs weekends


# calculate the average per time interval
t3 <- df_fill %>% group_by(weekday, interval) %>% summarise(records=n(), average_steps=mean(steps), total_steps=sum(steps))

# plot the lines using ggplot2
ggplot(t3, aes(x=interval, y=average_steps, color=weekday)) + geom_line() + facet_grid(weekday ~ .)


```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
