---
title: "Course 5 - Week 2 Assignment"
author: "Noel Figuera"
date: "May 15, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What is it required?

* Code for reading in the dataset and/or processing the data
* Histogram of the total number of steps taken each day
* Mean and median number of steps taken each day
* Time series plot of the average number of steps taken
* The 5-minute interval that, on average, contains the maximum number of steps
* Code to describe and show a strategy for imputing missing data
* Histogram of the total number of steps taken each day after missing values are imputed
* Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
* All of the R code needed to reproduce the results (numbers, plots, etc.) in the report

```{r }

library(ggplot2)
library(dplyr)

path <- 'activity.csv'
df <- read.csv(path,stringsAsFactors = FALSE)

# load the dataset

df$date <- as.Date(df$date,format="%Y-%m-%d")

# create a factor with two levels, weekday or weekend
df$weekday <- sapply(df$date, function(x) {
      if (weekdays(x) == "Sunday" || weekdays(x)=="Saturday") {
            return ('WEEKEND')
      } else {
            return ('WEEKDAY')
      }})

df$weekday <- as.factor(df$weekday)

```

# Analysis including missing values

## Histogram, mean and median

```{r }
# first analysis include all nas
df_na <- df[!is.na(df$steps),]

# total steps taken by day
s <- df_na %>% group_by(date) %>% summarise(records=n(), total_steps=sum(steps))

# create an histogram
ggplot(s, aes(total_steps)) + geom_histogram(bins=50)

# calculate mean and median
summary(s$total_steps)
```

## Interval plot excluding missing values

```{r }


# calculate the average per time interval
t <- df_na %>% group_by(interval) %>% summarise(records=n(), average_steps=mean(steps), total_steps=sum(steps))

# plot the lines using ggplot2
ggplot(t, aes(x=interval, y=average_steps)) + geom_line()
```

## Which time interval contains the highest on average?

830 minutes.

```{r }
# which on average contains the max?
head(t[order(-t$average_steps),], 1)
#835
```

## Define a strategy to complete missing cases

The strategy to follow is to use the average of the interval.

For that we use the match function. First, we asses the size of the issue:

```{r }
# analyse number of complete cases vs with nas
x <- c(length(df[!is.na(df$steps),]$steps), length(df[is.na(df$steps),]$steps))
names(x) <- c("Complete Cases", "With NAs")
barplot(x)
```

And after excuting the cleaning process we validate the results:

```{r }
# fill the na's with values using a simple strategy, substitute using the average of the 
# interval average
df_fill <- df
df_fill$steps[is.na(df$steps)] <- t$average_steps[match(df$interval,t$interval)][which(is.na(df$steps))]

# validate the results
x <- c(length(df_fill[!is.na(df_fill$steps),]$steps), length(df_fill[is.na(df_fill$steps),]$steps))
names(x) <- c("Complete Cases", "With NAs")
barplot(x)

```

## Histogram, mean and median

```{r }
##################################################################
#  We repeat the whole analysis for the dataset without NAs


# total steps taken by day
s2 <- df_fill %>% group_by(date) %>% summarise(records=n(), total_steps=sum(steps))

# create an histogram
ggplot(s2, aes(total_steps)) + geom_histogram(bins=50)

# calculate mean and median
summary(s2$total_steps)

```

## Interval plot excluding missing values

```{r }

# calculate the average per time interval
t2 <- df_fill %>% group_by(interval) %>% summarise(records=n(), average_steps=mean(steps), total_steps=sum(steps))

# plot the lines using ggplot2
ggplot(t2, aes(x=interval, y=average_steps)) + geom_line()

```

## Analyze weekdays & weekend pattern

```{r }
##################################################################
# Identify patterns on weekdays vs weekends


# calculate the average per time interval
t3 <- df_fill %>% group_by(weekday, interval) %>% summarise(records=n(), average_steps=mean(steps), total_steps=sum(steps))

# plot the lines using ggplot2
ggplot(t3, aes(x=interval, y=average_steps, color=weekday)) + geom_line() + facet_grid(weekday ~ .)


```
